---
title: "Cost-benefit analysis of a conservation policy under climate change: The Red Palm Weevil in Catalonia, Spain"

author:
  - name: Ángela Delgado Castillo
    affiliation: a
    footnote: "Corresponding Author: LuzAngela.Delgado@uab.cat"
  - name: Jeroen C.J.M.van den Bergh
    affiliation: a,b,c
  - name: Víctor Sarto i Monteys
    affiliation: a,d
address:
  - code: a
    address: Institute of Environmental Science and Technology, Universitat Autònoma de Barcelona, Edifici Z, 08193 Bellaterra, Spain.
  - code: b
    address: ICREA, Barcelona, Spain.
  - code: c
    address: Faculty of Economics and Business Administration, and Institute for Environmental Studies, VU University Amsterdam, The Netherlands.
  - code: d
    address: Departament d'Agricultura, Ramaderia, Pesca, Alimentació i Medi Natural Generalitat de Catalunya  Avinguda Meridiana, 38 08018 Barcelona (Spain).


abstract: |
   Invasive species are costly for human health, the environment and the economy. This burden is expected to rise. With limited budgets to address biological invasions, prioritisation is key for resource allocation. Recently, multiple frameworks have emerged to support this budget assignment, although it is not clear if current strategies are effectively integrating these frameworks. Amongst invasive species, insects are the costliest. In this article we evaluated a set of conservation policies in response to the arrival of the invasive beetle, the Red Palm Weevil (*Rhynchophorus ferrugineus*). The purpose of the selected schemes was to preserve non-native and non-threatened palm species exclusively used for ornamental purposes, in Catalonia, Spain. In a region with a large portion of land dedicated to agricultural activities, densely populated coastal areas with high volumes of trade, budgets to address biological invasions should be carefully assigned. Through a comprehensive cost-benefit analysis based on the total economic value framework, we found that the policies were not justified as the net social benefits were negative. In other words, the costs exceeded the benefits.

keywords: "Environmental valuation, *Phoenix. spp*, ecosystem services, *Rhynchophorus ferrugineus*."
  
date: "`r Sys.Date()`"
bookdown:
  fig_caption: yes
bibliography: library1.bib
header-includes:
   - \usepackage{caption,setspace,titlesec}
      \usepackage[section]{placeins}
      \usepackage[left=3cm, right=3cm, top=3cm]{geometry}
      \captionsetup[table]{font=normalsize,labelfont=normalsize}
      \captionsetup[figure]{font={normalsize,it},labelfont={normalsize,it}}
      \doublespacing
     
output: rticles::elsevier_article
---
\makeatletter
\def\ps@pprintTitle{%
\let\@oddhead\@empty
\let\@evenhead\@empty
\def\@oddfoot{\reset@font\hfil\thepage\hfil}
\let\@evenfoot\@oddfoot
}
\makeatother


```{r setup, include=FALSE}
x <-
  c("tidyverse",
    "officer",
    "bookdown",
    "fcuk",
    "knitr",
    "kableExtra",
    "kable",
    "formatR",
    "gridExtra",
    "jsonlite",
    "httr",
    "base64enc",
    "rlist",
    "RefManageR",
    "openxlsx",
    "BMA",
    "BAS",
    "statsr",
    "MASS",
    "deSolve",
    "FME",
    "RUnit",
    "plyr",
    "gdata",
    "citr",
    "FinancialMath",
    "quantmod",
    "lubridate")


lapply(x, require, character.only = TRUE)

opts_chunk$set(echo = FALSE)
options(knitr.table.format = "latex")
```

1. Introduction
==========================

The introduction of invasive alien species (IAS) can be considered as an important feature of the Anthropocene [@capinhaDispersalAlienSpecies2015; @lewisDefiningAnthropocene2015]. A worldwide growth in trade and transport for the past two centuries has increased the rate at which new IAS appear [@costelloUnintendedBiologicalInvasions2007; @pysekInvasiveSpeciesEnvironmental2010]. 1-16% of all species on Earth, depending on the taxonomic classification, qualify as potential alien species. For most taxonomic groups, the greater part of potential invasions has yet to occur [@seebensGlobalRiseEmerging2018].  

IAS often have a negative impact on human health, the economy and ecosystems
[@bradshawMassiveGrosslyUnderestimated2016; @colauttiCharacterisedProjectedCosts2006; @painiGlobalThreatAgriculture2016; @pimentelUpdateEnvironmentalEconomic2005; @vazquez-prokopecUnforeseenCostsCutting2010; @williamsEconomicCostInvasive2010].
Invasive insects (II) are likely to be the costliest species to humans with an annual minimum estimation of USD 70 billion globally [@bradshawMassiveGrosslyUnderestimated2016]. There are about 1.84 to 2.57 million insects species [@moraHowManySpecies2011] and to date, only an approximate 24% of potential insect invasions have taken place [@seebensGlobalRiseEmerging2018]. Depending on the policy implemented to lessen the impact of these invasions, the cost will be typically assumed by taxpayers, private entities, and residential property owners [@fenichelControlInvasiveSpecies2014; @funkBroadeningCaseInvasive2014; @lovettNonnativeForestInsects2016; @martelliEconomicImpactDengue2015; @vandammeOutofpocketHealthExpenditure2004].

To minimise costs borne by II, often the best policies are preventive which include measures in exporting countries and inspections of shipments at ports of entry [@lovettNonnativeForestInsects2016]. In instances of effective II introduction, governments can apply post-entry measures such as quarantines, surveillance and eradication programs [@kellerInternationalPolicyOptions2011; @thompsoncampbellFadingForestsNorth1994]. However, when budgets to address biological invasions are limited and invasions are rising, what should governments and individuals do? Prioritisation of which species, pathways, or sites to address is key [@mcgeochPrioritizingSpeciesPathways2016], as stated in Target 9 of the Aichi Biodiversity Targets of Convention on Biological Diversity [@conventiononbiologicaldiversityAichiBiodiversityTargets]. 

In this article we explore a set of conservation policies designed to address the arrival of an invasive insect. Our aim is to understand to what extent the crucial concept of prioritisation is being implemented in current policies. This assessment supports policy makers at a local, regional and international level to improve resource allocation by considering the environmental and possibly socio-economic impacts [@blackburnUnifiedClassificationAlien2014; @kumschickConceptualFrameworkPrioritization2012] and selecting the actions that best avoid or lessen the impact of invasive species [@mcgeochPrioritizingSpeciesPathways2016].

The conservation policies selected [^1] were designed to preserve two species: Phoenix dactylifera (the Date palm) and Phoenix canariensis (the Canary palm), both non-native and considered of 'least concern' by the IUCN Global Species Programme Red List Unit [@beechPhoenixCanariensisIUCN2017] but susceptible to an II in Catalonia, Spain. These two species have come under pressure in the region of study due to the arrival of the red palm weevil (*Rhynchophorus ferrugineus*, Coleoptera: Dryophthoridae) or as it is known locally, the ‘morrut de les palmeres’ (Catalan for palm weevil). II like the red palm weevil (RPW) are introduced through trade and transport [@costelloUnintendedBiologicalInvasions2007; @margolisHowTradePolitics2005; @meyersonInvasiveAlienSpecies2007]. Merchandise imports [@levineForecastingBiologicalInvasions2003] and economic and demographic variables have been found to influence invasion levels [@pysekDisentanglingRoleEnvironmental2010].

In Catalonia, Spain, there are over 800 gardens, nurseries and plant traders combined, 24% of them trading with palm trees. Catalonia has five administrative regions, all of them trading with palms which help increase risk for RPW expansion (Table \ref{tab:nurseries}). Sites with large volumes of trade and transport and adjacent to high human population densities such as Catalonia, likely benefit from prioritizing invasions [@mcgeochPrioritizingSpeciesPathways2016].

[^1]: The selected policies are Ordre ARP/343/2006, Ordre AAR/226/2009, Ordre AAR/2802/2010, and Ordre AAM/56/2011. These are explained in further detail in the Appendix A.

```{r nurseries table, echo=FALSE}
options(scipen = 999, digits = 2)
all_nurseries <- c(348, 162, 148, 98, 106, 862)
palm_nurseries <- c(59, 63, 10, 36, 37, 205)
region <-  c("Barcelona",
             "Girona",
             "Lleida",
             "Tarragona",
             "Terres de l'Ebre",
             "Total")
df_nurseries <- data_frame(region,
                           palm_nurseries,
                           all_nurseries)
df_nurseries$region <- as.factor(df_nurseries$region)
colnames(df_nurseries) <- c("Region",
                            "With palms",
                            "All")
df_nurseries <- df_nurseries %>%
  mutate(Ratio = `With palms` / All)
kable(
  df_nurseries,
  caption = "\\label{tab:nurseries}Gardens, nurseries and plant traders in Catalonia per Administrative Region.",
  booktabs = T,
  escape = F
) %>%
  kable_styling(latex_options = c("hold_position"))%>%
  footnote(general="GENCAT, 2018.",
           general_title = "Source: ",
                     footnote_as_chunk = T)
```


The selected policies aim to mitigate the insect's impact on susceptible palm species. We will use a cost-benefit analysis as a tool to assess the selected policies since it has recently been highlighted to overcome "pitfalls" present in other methods: (1) prioritisations are sometimes based on expert judgements and scores are debatable, (2) summation of the scores is tricky, (3) inclusion of management expenditure is not usual and (4) relationships between species are often excluded [@courtoisCostBenefitApproach2018].

1.1 The RPW around the world and in Catalonia
==========================

The RPW is native to Southeast Asia and Polynesia [@wattanapongsiriRevisionGeneraRhynchophorus1966] being essentially a pest of palms *(Arecaceae)*, with 29 species listed as host plants [@cabiDatasheetRhynchophorusFerrugineus2017]. It has also been mentioned from sugarcane, Saccharum officinarum, *(Poaceae)* and sentry plant or maguey, *Agave americana (Agavaceae)* [@cabiDatasheetRhynchophorusFerrugineus2017] although this would require confirmation. It was introduced in other parts of the world mainly through import of palm trees or offshoots from native areas. For instance, it reached Egypt in 1992 from the United Arab Emirates, and Spain in 1995 from imported egyptian palms [@ferryRedPalmWeevil2002].


The RPW lays eggs on the crown of palm trees. Larvae then penetrate it eventually reaching the crown core where the meristematic tissue resides. Subsequent feeding on this tissue will eventually kill the palm tree. Because the whole larval stage occurs within the palm tree, it is concealed from plain eyesight and hence hard to detect. 

In Spain, palm damage by this beetle was first observed in 1993 on the coastal front of the Granada province [@barrancoNuevoCurculionidoTropical1996]. In 2004 it reached the Autonomous Community of Valencia [@eppoRhynchophorusFerrugineus2008]. It was officially declared  a pest in Catalonia in 2006 [@generalitatdecatalunyaOrdreARP3412006] where it presumable arrived between 2003 and 2004 to the Ebro river delta and Tarragona region (South Catalonia).

The Catalan Department of Agriculture has made efforts to identify in what regions (“comarques”) the beetle is present, resulting in the pattern shown in . This illustration, provided by GENCAT, colors the areas where the RPW has been reported upto November 2016. The colored area is along the coast since this is where most palm trees are present, although a few inner regions are also included. 

Figure \ref{fig:catalonia} shows the invasion status of RPW in Catalonia up to November 2016 [@generalitatdecatalunyaRedPalmWeevil2017].

\begin{figure}[!htb]
\centerline{\includegraphics[width=20pc]{2016invasion.png}} 
\caption{2016 invasion status. Source:[@generalitatdecatalunyaRedPalmWeevil2017]}
\label{fig:catalonia} 
\end{figure}

Some areas in the Mediterranean have also been colonised. Sicily, for instance, was reached in 2005 where palm damage was categorised as a threat to cultural heritage [@manachiniExoticInsectPests2013; @periRedPalmWeevil2013]. Other countries where it appeared for the first time in this period include Turkey (2005), Cyprus (2006), Greece (2006) and Malta (2007) [@mizziRedPalmWeevil2009; @yuezhongRedPalmWeevil2009].
The beetle has also invaded other parts of the world. In Japan, it was first observed in 1975 in the island of Okinawa, and later (in 1998) spread north to the island of Kyushu [@abeLifeHistoryRed2009]. During the 1990s, the RPW was intercepted on several occasions in China. In 2007, several cases in the Zhejiang province were detected signalling a spread throughout Southern China [@yuezhongRedPalmWeevil2009]. Other Asian countries where the RPW is present include: India, Pakistan, Sri Lanka, Myanmar, Indonesia, the Philippines, and the Gulf states [@abeLifeHistoryRed2009]. In the Americas, the RPW was detected in the Caribbean islands of Aruba and Curaçao in 2009 which meant a risk of the weevil being exported to South America [@rodaRedPalmWeevil2011]. On October 2010, the USDA’s Animal and Plant Health Inspection Service (APHIS) announced the arrival of the closely related species *Rhynchophorus vulneratus* to the city of Laguna Beach, California. Today, the RPW is present in over 70 countries [@cabiDatasheetRhynchophorusFerrugineus2017]. 


2. Model
============

2.1 Cost-benefit analysis
============
We will perform a cost-benefit analysis assessing a set of conservation policies put into place to protect susceptible species from the RPW. Based on the policies selected, two species appear: *Phoenix canariensis*, from here on, canary palm, and *Phoenix dactylifera*, from here on date palm [@generalitatdecatalunyaORDERARP3432006]. The former was reported in 99.69% and the latter in 0.23% of RPW cases in the region [@generalitatdecatalunyaPrevencioLluitaContra2011]. Although date palms currently represent a small fraction of reports, we include them because in absence of canary palms, the RPW would likely shift to infesting date palms. 
The first step of a cost benefit analysis is to define what constitute as costs and benefits. We interpret as costs all expenditures with the direct aim of protecting and preserving palm trees in presence of the RPW as a direct output of the conservation policies set into place. The following activities fall into the management category and include: prevention and inspection of healthy palms, treatment of infested palms and removal of palms that are too infested to be salvaged. However, a full assessment should include not only management expenditure, but also, government expenditure associated to survey, research, regulation, management and outreach [@aukemaEconomicImpactsNonNative2011]. To better understand the rationale behind the accouting used in this study, a few points are made on the policies. 
The main variation in terms of the funding agent is that in 2011, inspection, prevention, treatment and removal costs were transfered from the Catalonian goverment (GENCAT), to municipalities and private owners. From 2005 to 2011, GenCat was responsible for research and outreach in addition to the management activities described (excluding replanting). From 2011, onwards, municipalities and property owners are responsible for management activities while GenCat continues funding research and outreach for the pest. Although replanting is not explicit in the Orders, between 2005 to present, replanting has not been null albeit total stock or rates unknown [@congostrinaBarcelonaPlantaPalmeras2017; @puricaroBarcelonaAtesora242017], we therefore include replanting on behalf of municipalities and property owners throughout the entire period.

To define benefits, we will use the total economic value (TEV) framework [@ledouxValuingOceanCoastal2002]. The TEV framework classifies the value of an asset into two main categories: use and non-use values.The main difference between these values is that use values are related to the use of the asset while non-use values relate to the motivation for conserving the asset for future generations or for its own sake. Figure \ref {fig:tev} further illustrates this point by listing indirect and direct values derived from palm trees. The former includes aesthetic benefits and carbon sequestration. Although in some places palm trees have direct values (i.e. cultivation), as will be described in the following paragraphs, in the scope of this article, no direct values were identified. 

\begin{figure}[!htb]
\centerline{\includegraphics[width=30pc]{tev.png}} 
\caption{Total economic value framework.}
\label{fig:tev}
\end{figure}

Canary and date palms are non-native to the Iberian Peninsula. They were introduced in ancient times as evidenced by depictions of late Iberian pottery and coinage dating back to the third to first century BC [@al-khayriDatePalmGenetic2015]. Although canary palms produce edible fruits they are not used in agriculture, as opposed to date palms. In Spain, the land area dedicated to date cultivation is marginal with an average of ~600 hectares since 1961 to present [@faostatFAO2018]. In Catalonia, our region of study, date cultivation is virtually non-existent. Palms are planted in urban areas, resorts, public spaces and private gardens mainly near the coast [@al-khayriDatePalmGenetic2015]. Currently, they are classified as as species of "least concern" on the IUCN Red List of Threatened Species [@beechPhoenixCanariensisIUCN2017]. This means that they do not qualify as threatened, near threatened or conservation dependent before 2001.

Given this context and its almost exclusive ornamental use, we define benefits as follows. We identify two indirect use values as benefits: carbon sequestration and enhanced aesthetic value, the latter being the main reason behind palm tree availability in Catalonia. 


2.2 A model of tree stocks under RPW infestation
============

We modelled the evolution of tree stocks over time as a system dynamics problem since this approach allows us to understand nonlinear behavior over time using stocks, flows and feedback loops [@forresterBeginningSystemDynamics1989]. The two palm species have similar processes with different initialisation values so we implemented two separate modules. From this point onward, susceptible palm species will be referred to as ‘tree’. The model was designed with three stocks: healthy, infested and treated trees. It is graphically illustrated with Figure \ref{fig:model}, created using Vensim software.

\begin{figure}[!htb]\centerline{\includegraphics[width=30pc]{model.png}} \caption{Model representation.}\label{fig:model}\end{figure}

The total amount of trees at time $S_t$ will be the sum of healthy trees, $S_t^H$, the sum of infested trees, $S_t^I$, and the sum of trees in treatment, $S_t^T$.

\begin{equation} \label{eq:total_stock}
S_t=S_t^H+S_t^I+S_t^T
\end{equation}

Healthy trees, $S_t^H$, are all susceptible trees with two incoming flows: replanted trees, $F_{t,t-1}^R$, and successfully treated trees, $F_{t,t-1}^{ST}$. Prevention consists on a series of chemical applications to protect the tree from the RPW. Outgoing flows are trees in which prevention has failed, $F_{t,t-1}^{FP}$. Therefore, the equation for the healthy net flow, $S_t^H-S_{t-1}^H$, between $t-1$ and $t$ is:

\begin{equation} \label{eq:stock_healthy}
S_t^H-S_{t-1}^H=F_{t,t-1}^R+F_{t,t-1}^{ST}-F_{t,t-1}^{FP}
\end{equation}

The outflow, $F_{t,t-1}^{FP}$, depends on the force of infestation, denoted by lambda ($\lambda_t$). The force of infestation is the rate at which susceptible trees become infested per unit time. It is adapted from the force of infection [@vynnyckyIntroductionInfectiousDisease2010]. The rate of infestation is related to the number of infested trees, $S_t^I$, the more trees are infested, the more likely more trees will become infested and to the average number of trees that are planted next to each other, denoted with Beta ($\beta_t$). 

Thus, the force of infestation, $\lambda_t$, at time $t$ is:

\begin{equation} \label{eq:lambda}
\lambda_t=\beta_t*S_t^I
\end{equation}						 

Beta ($\beta_t$) in turn, depends on the effective contact rate ($C_E$), defined as the contact rate sufficient to lead to transmission if it occurs between an infested and a susceptible tree [@vynnyckyIntroductionInfectiousDisease2010] with units $\frac{palms}{(palm/year)}$ and on the total stock, $S_t$, at time $t$.

\begin{equation} \label{eq:beta}
\beta_t=\frac{C_E}{S_t}		
\end{equation}	

The failed prevention outflow, $F_{t,t-1}^{FP}$, is the outflow of trees in which prevention has failed ($FP$) product of the number of healthy trees,$S_t^H$, and lambda, $\lambda_t$, the force of infection at time $t$.

\begin{equation} \label{eq: failed prevention}
F_{t,t-1}^{FP}=S_t^H*\lambda_t
\end{equation}

Failed prevention, $F_{t,t-1}^{FP}$, is an incoming flow to the infested tree stock, $S_t^I$. Infested trees are either treated, $F_{t,t-1}^{TR}$, or not treated, $F_{t,t-1}^{UTR}$, leading to two outgoing flows for this stock. Therefore, the equation for the net infested flow between $t-1$ and $t$ is:

\begin{equation} \label{eq: stock_infested}
S_t^I-S_{t-1}^I=F_{t,t-1}^{FP}-F_{t,t-1}^{UTR}-F_{t,t-1}^{TR}
\end{equation}

Both treated, $F_{t,t-1}^{TR}$, and untreated flows, $F_{t,t-1}^{UTR}$ depend on the probability, ($\tau$), of being treated. Thus,

\begin{equation} \label{eq: treated_prob}
F_{t,t-1}^{TR}=S_{t-1}^I*\tau
\end{equation}

\begin{equation} \label{eq: untreated_prob}
F_{t,t-1}^{UTR}=S_{t-1}^I-F_{t,t-1}^{TR}
\end{equation}

Treated trees, $F_{t,t-1}^{TR}$,is an incoming flow to treated trees, $S_t^T$. Treated trees has outgoing flows representing either failed, $F_{t,t-1}^{FT}$, or successful treatment, $F_{t,t-1}^{ST}$. Thus,

\begin{equation} \label{eq: stock_treated}
S_t^T-S_{t-1}^T=F_{t,t-1}^{TR}-F_{t,t-1}^{FT}-F_{t,t-1}^{ST}
\end{equation}

Both successful, $F_{t,t-1}^{ST}$, and failed treatment flows, $F_{t,t-1}^{FT}$, depend on the probability of successful treatment sigma ($\sigma$). Thus,

\begin{equation} \label{eq: successful_treatment}
F_{t,t-1}^{ST}=S_t^T*\sigma
\end{equation}
\begin{equation} \label{eq: failed_treatment}
F_{t,t-1}^{FT}=S_t^T-F_{t,t-1}^{ST}
\end{equation}

Finally, to keep track of trees that were lost each year, $S_t^L$, due to lack of treatment, $F_{t,t-1}^{UTR}$ or failed treatment, $F_{t,t-1}^{FT}$, we define the lost stock, $S_t^L-S_{t-1}^L$, between $t-1$ and $t$ as:

\begin{equation} \label{eq: lost_stock}
S_t^L-S_{t-1}^L=F_{t,t-1}^{UTR}+F_{t,t-1}^{FT}
\end{equation}

Similarly to keep track of trees that have been replanted each year, $S_t^R$, due to the incoming flow of replanted trees, $F_{t,t-1}^R$, we define the stock of replanted trees $S_t^R-S_{t-1}^R$, between $t-1$ and $t$ as:

\begin{equation} \label{eq: replanted_stock}
S_t^R-S_{t-1}^R=F_{t,t-1}^R
\end{equation}

Costs
=======

We now introduce equations to calculate the costs. These are divided into two main groups. The first is management costs assumed by municipalities and private palm owners from 2011 onwards. These management costs, $M_t$, for trees in year $t$ are described as the sum of inspection, $C_t^{inspect}$, prevention, $C_t^{prevent}$, treatment costs, $C_t^{treat}$, replanting, $C_t^{replant}$, and removal costs, $C_t^{remove}$. The second group is GenCat costs which includes management costs (except replanting) but also additional activities such as research and outreach. These were provided from 2005 to 2017 and since no specific breakdown was provided by GenCat so they are added as a lump sum.

\begin{equation} \label{eq:management_costs} 
M_t=C_t^{inspect}+C_t^{prevent}+C_t^{treat}+C_t^{replant}+C_t^{removal}
\end{equation} 
    	
Inspection costs, $C_t^{inspect}$, are the costs of inspecting healthy trees in year $t$.

\begin{equation} \label{eq:inspection_costs} 
C_t^{inspect}=S_t^H*c^{inspect}
 \end{equation} 						
 								 
Here, $S_t^H$, is the number of healthy trees ($H$)in year $t$ and $c^{inspect}$ is the inspection unit cost. 
Prevention costs, $C_t^{prevent}$, are the costs of applying preventative treatments to healthy trees in year $t$.

\begin{equation} \label{eq:prevention_costs} 
C_t^{prevent}=S_t^H*c^{prevent}
\end{equation} 
  												
Here, $S_t^H$ is the number of healthy trees ($H$)in year $t$ and $c^{prevent}$ is the prevention unit cost. 

Treatment costs, $C_t^{treat}$, are the costs of treating a tree selected for treatment ($T$) to improve its health in year $t$.

\begin{equation} \label{eq:treatment_costs} 
C_t^{treat}=S_t^T*c^{treat}
\end{equation} 									

Here, $S_t^T$ is the number of treated trees ($T$) in year $t$ and $c^{treat}$ is the treatment unit cost. 

Replanting costs, $C_t^{replant}$, are the costs of replanting trees in year $t$.

\begin{equation} \label{eq:replanting_costs} 
C_t^{replant}=S_t^R*c^{replant}
\end{equation} 	

Here, $S_t^R$, is the number of replanted trees ($R$) in year $t$ and $c^{replant}$ is the unit cost of replanting a new tree. 

$C_t^{remove}$, corresponds to the costs of permanently removing a tree in year $t$, which the RPW has damaged beyond the point of treatment, $S_t^L$.

\begin{equation} \label{eq:removal_costs} 
C_t^{remove}=S_t^L*c^{remove}
\end{equation} 						     	

Here, $S_t^L$ is the number of lost trees in year $t$ and $c^{remove}$ is the unit cost of removing a lost tree.


Benefits of palm trees
==========
As mentioned in the introduction, trees provide different benefits ($VB$). We will consider two benefits provided by palm trees in Catalonia (Eq. \ref{eq:benefits}), (1) those derived from carbon sequestration and (2) those derived from forestry adjacency or in this instance, palm tree adjacency.

\begin{equation} 
\label{eq:benefits}VB=VCS+VPA
\end{equation}

Where, $VCS$  is the value derived from carbon sequestration by palm trees in Catalonia and $VPA$  is the value derived from palm tree adjacency. The benefits of carbon sequestration, $VCS$, will be calculated using the social cost of carbon dioxide. The social cost of carbon ($SCC$) is an economic indicator that represents the net impacts from the global climate change that result from a one tonne increase in carbon dioxide emissions [@iawgTechnicalUpdateSocial2016].

\begin{equation} 
\label{eq:social_cost}VCS=SCC*c_s*S_t 
\end{equation}

					           	    	
Where, $SCC$ is the social cost of carbon dioxide, $c_s$ is the amount of carbon sequestered by a tree; and $S_t$ is the total stock in year $t$.

Benefits of tree adjacency can be captured by using the hedonic property-value method. This method helps us estimate the value that trees contribute to the sale value of homes. Several studies have focused on quantifying forest adjacency (Donovan & Butry, 2010; Sander et al., 2010; Tyrväinen & Miettinen, 2000) with results stating that trees contribute between 0.29 to 5% of property values of private residences. In this article we attempt to capture the value that, specifically, palm trees contribute to property prices. 

We assume that the total value of a property in Catalonian municipalities, where trees are viable, includes a component of palm trees in addition to other market effects. This component of forestry adjacency is captured in property values, $PV$:

\begin{equation} 
\label{eq:property_values}PV=MP+PA
\end{equation}
						
Where, $MP$ is the market price of a property without adjacent palm trees and $PA$ is the value derived from palm tree adjacency.

Percentage changes due to palm tree adjacency, $\%PA$, are linked to the percentage variation in property values associated with forestry adjacency and to the number of palm trees close to a property (Eq. \ref{eq:palm_adjacency})

\begin{equation} 
\label{eq:palm_adjacency}\%PA=-\gamma*\frac{S_{2017}-S_{2005}}{S_{2005}}
\end{equation}

Here, $\gamma$, is the percentage variation in property values associated to forestry adjacency and $S_{2005}$ and $S_{2017}$ are the estimated number of palm trees in 2005 and 2017, respectively. The difference in a property’s property value $PV$ between 2005 and 2017 is derived from three changes: market price ($MP$), inflation ($I$), and palm tree adjacency ($PA$).

\begin{equation} 
\label{eq:pv_2017}PV_{2017}-PV_{2005}=PV_{2005}*(\%MP+\%I+\%PA)
\end{equation}

Here, $\%MP$, is the percentage change in property value, $\%I$, is the percentage change in inflation, and $\%PA$,  is the percentage change is palm tree adjacency between 2017 and 2005. Solving for $PV_{2005}$ yields Eq. \ref{eq:pv_2016}.

\begin{equation} 
\label{eq:pv_2016}PV_{2005}=\frac{PV_{2017}}{1+\%MP+\%I+\%PA}
\end{equation}    					 

If $\%MP$ and $\%I$ are zero, the property value loss associated to the beetle’s impact on palm tree adjacency in Catalonia between 2017 and 2005 is summarised in Eq. \ref{eq:pv_diff}.

\begin{equation} 
\label{eq:pv_diff}PV_{2017}-PV_{2005}=PV_{2017}-\frac{PV_{2017}}{1+\%PA}=PV_{2017}*\left(1-\frac{1}{1+\%PA}\right)
\end{equation}   

Total costs and benefits 
=========

The net social benefits (NSB) of the current management policy against the RPW in Catalonia is the difference between benefits and costs brought to present value (Eq. \ref{eq:nsb}).

\begin{equation} 
\label{eq:nsb} NSB=\sum_{t}\frac{B_{t}-C_{t}}{(1-r)^t}
\end{equation}

Here, $B_{t}$ denotes the monetary benefits of the policy, $C_{t}$, management costs and, $r$, the discount rate in year $t$. 
						

2.3 Estimation of Parameter Values
========================

As explained, the model has three parameters: effective contact rate ($C_E$), probability of being treated ($\tau$), and probability of successful treatment ($\sigma$). In this section, the process to estimate each parameter is explained.

\subsubsection{Effective contact rate ($C_E$)}

Effective contact rate, $C_E$, is defined as the contact rate sufficient to lead to transmission if it occurs between an infested and a susceptible tree [@vynnyckyIntroductionInfectiousDisease2010] with units $\frac{palms}{(palm/year)}$ 
It can take up values [1,$\infty$] to which we assigned [1,2] since a single infested tree can lead to infestation of a susceptible tree within a certain radius.

\subsubsection{Probability of being treated ($\tau$)}

Interviews with three government officials and two local independent firms offering pest management services (inspection, prevention, treatment, and removal) were conducted with the goal of eliciting treatment success probabilities. Parameter ($\tau$) has been defined as the probability that an affected tree is treated. $\tau$ can take values between [0,1] to which we assigned the interval [0.5, 1]. Since legislation states that treatment of all affected trees is mandatory, we have assigned the upper boundary as 1 for the case in which 100% of affected trees are treated. According to interviewees, there are instances where treatment of affected trees does not occur. This may happen deliberately or accidentally. Furthermore, no official data exists describing the behaviour of private owners while municipality reports are compromised since they must reflect compliance with the law, which means treatment in all cases. In order to account for this unknown behaviour where treatment doesn't occurr but is unknown, the lower boundary was set to a probability of 0.5 (i.e. coin toss).

\subsubsection{Probability of successful treatment ($\sigma$)}

Sigma ($\sigma$) is the probability that treatment is successful. Again, as it is a probability, it can take values between [0,1]. Success is defined as a tree being sufficiently recovered to rejoin the healthy stock. Interviewees were consistent in their responses stating that once a tree becomes infested, survival probability is slim to none. For this reason, the interval [0,0.5] was assigned as it reflects the pessimistic outlook on infested trees under treatment from null recovery rates up to 0.5 (i.e. coin toss).


2.4 Model calibration
========================

We implemented the model using the deSolve package in R Software [@soetaertSolvingDifferentialEquations2010]. To run the model, however, two initial stock values had to be defined for $t=2005$, healthy, $S_{2005}^H$, and infested, $S_{2005}^I$, stock. In this section we explain how these two initialisation values were set. 

Healthy stock 2005
========

Since public tree inventory is decentralised and private owners have agency to change their landscape, public and private tree records were unavailable. Instead, we estimated tree stocks. To estimate total stock in 2005 (Eq. \ref{eq:2005stock}), $S_{2005}$, we first estimated values for 2017, $S_{2017}$, and added the minimum loss of stock during the invasion period, $S_{2017-2005}^L$. The minimum loss was provided by the regional government. The process for estimating stock in 2017 in will be now explained with more detail. 

\begin{equation} 
\label{eq:2005stock}S_{2005}=S_{2017}+S_{2017-2005}^L
\end{equation}

As previously mentioned, the availability of both palm trees in the region of study is human-mediated given its almost exclusive ornamental use. Due to this management of palm trees in the region of study, we identified multiple human-centred factors as explanatory variables (namely population, GDP per sector and municipality area). Furthermore, we consulted with two experts in forestry and ornamental plants who suggested distance to the Western Mediterranean Basin plays a key role in the presence of palm trees.

Data for aforementioned variables was retrieved from the Catalonian statistical department (IDESCAT) for all municipalities. Distance was calculated using polygon centroids. GDP per sector was not available for all 947 municipalities, however, it was available for all 'comarcas', which are a higher administrative level made up of 42 divisions. 'Comarcas' GDP per sector was used for municipalities lacking these data. 

We selected a stratified sampling strategy to collect data on the both species. This strategy was selected because we wanted all comarcas to be represented within the sample. We set out to collect a sample size of 50 municipalities. The total number of municipalities is 947 and they are divided in 42 comarcas, which we defined as the strata. For each strata, we calculated the sample size as: $\frac{Sample\ Size}{Total\ population}*Strata\ size$. For example, for the comarca Baix Camp with 28 municipalities, calculations were: $\frac{50}{947}*28=1.48\approx 2\ municipalities$. Repeating this process for all comarcas and rounding decimals upwards yielded 73 municipalities. 

We collected data on tree stocks using static maps from Google Maps. The following procedure was done for each of the 73 municipalities. First, using a shapefile for the region and the R Packages sf and GGMap [@kahleGgmapSpatialVisualization2013], a grid for each municipality was produced where each cell in the grid represented a separate image. Examples of the grids are shown in Figure \ref{fig:grids} where the upper row is the municipality polygon layered on to a satellite image and the lower row is the corresponding grid. Over 25,000 images were individually assessed for palms trees of each species. 

\begin{figure}[!htb]
\centerline{\includegraphics[width=30pc]{grids.png}} 
\caption{Satellite pictures of municipality polygons (upper row), municipality grids for palm tree estimation (lower row).}
\label{fig:grids} 
\end{figure}

Once observations were collected, we ran a linear model under Bayes model averaging. At this point we had no criteria to select one model over another so Bayes model averaging was useful as this model averages multiple models according to their posterior probability. The model was specified as follows: all seven explanatory variables were included, a uniform distribution was asigned as the model prior, and the Bayes Information Criterion ($BIC$)[^2] was used as the prior distribution for regression coefficients since we were in a non-informative situation. 

```{r reading collected data, include=FALSE}
#Read palm data
palms <- read.xlsx("palm_counts.xlsx")
palms <- palms[1:73,] #rows 1 through 73
#Create ser, ag, indu and cons variables.
#Aggregate by ownership
palms <- palms %>%
  mutate(canary=canary_priv+canary_pub,
         date=date_priv+date_pub,
         ag=Agriculture/Total,
         indu=Industry/Total,
         cons=Construction/Total,
         ser=Services/Total)

#Read spreadsheet for all municipalities
final<- read.xlsx("final.xlsx")
#Create ser, ag, indu and cons variables.
final <- final %>%
    mutate(
         ag=Agriculture/Total,
         indu=Industry/Total,
         cons=Construction/Total,
         ser=Services/Total)
```

```{r Looking at distributions, include=FALSE}
boxplot_date<- ggplot(palms,aes(x=,y=date))+
  geom_boxplot()+
  coord_flip()

boxplot_canary<- ggplot(palms,aes(x=,y=canary))+
  geom_boxplot()+
  coord_flip()
```

```{r ape, include=FALSE }
set.seed(42)
n <-  nrow(palms)
n_cv <-  50
ape_date <-  matrix(NA, ncol=4, nrow=n_cv)
colnames(ape_date) = c("BMA", "BPM", "HPM", "MPM")

for (i in 1:n_cv) {
  train = sample(1:n, size=round(.90*n), replace=FALSE)
  date_train = palms[train,]
  date_test = palms[-train,]
  
  bma_train_date = bas.lm(date ~ ag+ser+cons+indu+distance+population+area, data=date_train, 
                          prior="BIC", modelprior=uniform(), initprobs="eplogp")
  yhat_bma = predict(bma_train_date, date_test, estimator="BMA")$fit
  yhat_hpm = predict(bma_train_date, date_test, estimator="HPM")$fit
  yhat_mpm = predict(bma_train_date, date_test, estimator="MPM")$fit
  yhat_bpm = predict(bma_train_date, date_test, estimator="BPM")$fit
  ape_date[i, "BMA"] = cv.summary.bas(yhat_bma, date_test$date)
  ape_date[i, "BPM"] = cv.summary.bas(yhat_bpm, date_test$date)
  ape_date[i, "HPM"] = cv.summary.bas(yhat_hpm, date_test$date)
  ape_date[i, "MPM"] = cv.summary.bas(yhat_mpm, date_test$date)
}

boxplot(ape_date)

apply(ape_date, 2, mean)

set.seed(42)
n <-  nrow(palms)
n_cv <-  50
ape_canary <-  matrix(NA, ncol=4, nrow=n_cv)
colnames(ape_canary) = c("BMA", "BPM", "HPM", "MPM")

for (i in 1:n_cv) {
  train = sample(1:n, size=round(.90*n), replace=FALSE)
  canary_train = palms[train,]
  canary_test = palms[-train,]
  
  bma_train_canary = bas.lm(canary ~ ag+ser+cons+indu+distance+area+population, data=canary_train, 
                          prior="BIC", modelprior=uniform(), initprobs="eplogp")
  yhat_bma = predict(bma_train_canary, canary_test, estimator="BMA")$fit
  yhat_hpm = predict(bma_train_canary, canary_test, estimator="HPM")$fit
  yhat_mpm = predict(bma_train_canary, canary_test, estimator="MPM")$fit
  yhat_bpm = predict(bma_train_canary, canary_test, estimator="BPM")$fit
  ape_canary[i, "BMA"] = cv.summary.bas(yhat_bma, canary_test$canary)
  ape_canary[i, "BPM"] = cv.summary.bas(yhat_bpm, canary_test$canary)
  ape_canary[i, "HPM"] = cv.summary.bas(yhat_hpm, canary_test$canary)
  ape_canary[i, "MPM"] = cv.summary.bas(yhat_mpm, canary_test$canary)
}

boxplot(ape_canary)

apply(ape_canary, 2, mean)
```

During the model selection stage, we considered several estimators: best predictive model (BPM), median probability model (MPM) and, highest probability model (HPM). In order to select amongst estimators, we computed the average predictive error (APE) for each one. BPM, the estimator that has predictions closest to BMA and corresponding posterior standard deviations, was selected for date palms and HPM, the model with the lowest $BIC$ was selected for canary palms. In terms of coefficients this meant, distance, population, fraction of GDP from agriculture and industry, for date palms; and distance, population, and fraction of GDP from industry, for canary palms.

```{r coefficients, include=FALSE}
BMA_date <-  bas.lm(date ~ ag+ser+cons+indu+distance+population+area, data = palms,
                      prior = "BIC",
                      modelprior = uniform())
BPM_pred_date =  predict(BMA_date, estimator="BPM", se.fit=TRUE)
BMA_date$namesx[BPM_pred_date$bestmodel+1]

lm_canary = lm(canary ~ ag+ser+cons+indu+distance+population+area, data=palms)

BMA_canary <-  bas.lm(canary ~ ag+ser+cons+indu+distance+population+area, data = palms,
                      prior = "BIC",
                      modelprior = uniform())
n <- length(BMA_canary)
stepAIC(object=lm_canary, k =log(n))
```

[^2]:$BIC= -2*log(likelihood)+log(n)*\#parameters$, where $n$ is the sample size. We will have a tradeoff between the goodness of fit (the extent to which observed data matches the values expected by theory) $n*log(1-R^2)$ and the model complexity represented with $log(n)*\#parameters$. We choose the model that has the smallest $BIC$. 

```{r tree models, include=FALSE}
lm_date <- lm(date ~ indu+distance+population+ag, data=palms)
 
lm_canary <- lm(canary ~ indu+distance+population, data=palms)
```

```{r date predictions, include=FALSE}

date_BMA_prediction<-predict(lm_date,newdata= final) #Apply model to new date
date_BMA_df<- as_tibble(date_BMA_prediction) 
names(date_BMA_df) <- c("date") 
date_BMA_df%>%
  mutate(date=if_else(date>0,date,0) #Remove negatives 
         )->date_BMA_df #Store in new dataframe
test_final <- rownames_to_column(final,var="index") #Adding id to new data
test_interval <- rownames_to_column(date_BMA_df,var="index") #Adding id to predictions
test_date<- test_final %>%
 left_join(test_interval,by="index") #Joining predictions with more info
```

```{r canary predictions, include=FALSE}

canary_BMA_prediction<-predict(lm_canary,newdata= final) #Apply model to new canary
canary_BMA_df<- as_tibble(canary_BMA_prediction) 
names(canary_BMA_df) <- c("canary") 
canary_BMA_df%>%
  mutate(canary=if_else(canary>0,canary,0) #Remove negatives 
  )->canary_BMA_df #Store in new dataframe
test_final <- rownames_to_column(final,var="index") #Adding id to new data
test_interval <- rownames_to_column(canary_BMA_df,var="index") #Adding id to predictions
test_canary<- test_final %>%
  left_join(test_interval,by="index") #Joining predictions with more info
```

```{r predictions summary, include=FALSE}
#Count how many trees we have of each
test_date%>%
  summarise(sum=ceiling(sum(date)),
            sum_2017=ceiling(sum(date)))->sum_date
test_canary%>%
  summarise(sum=ceiling(sum(canary)),
            sum_2017=ceiling(sum(canary)))->sum_canary
```

```{r prediction plots, include=FALSE}
date_mean <- ggplot(test_date, aes(
 x = distance,
  y = date,
color = cut(indu, breaks = 4),
    size= population)) +
  geom_jitter(alpha=0.5) +
  geom_text(
    aes(x = test_date$distance, y = test_date$date),
    label = test_date$Nom,
    size = 5,
    check_overlap = TRUE,
    inherit.aes = FALSE,
    hjust = 0,
    nudge_x = 0.005
  ) +
  theme_classic() +
  guides(
    size = guide_legend(title = "Population"),
      color=guide_legend(title = "Fraction of GDP from industry")
  ) +
  labs(
    title= paste("Date palm trees 2017,","n =",sum_date$sum_2017,sep = " "),
   subtitle = "Vs. distance, population and fraction of GDP from services.",
    x = "Distance to coastline (km)",
    y = "Date palms"
  )+ theme(
        text = element_text(size=20))
ggsave(
  "Date_palms.png",
  width = 7,
  height = 8,
  bg = "transparent"
)

canary_mean <- ggplot(test_canary, aes(
  x = distance,
  y = canary,
color = cut(indu, breaks = 4),
  size = population)
) +
  geom_jitter(alpha=0.5) +
  geom_text(
    aes(x = test_canary$distance, y = test_canary$canary),
    label = test_canary$Nom,
    size = 5,
    check_overlap = TRUE,
    inherit.aes = FALSE,
    hjust = 0,
    nudge_x = 0.005
  ) +
  theme_classic() +
  guides(
    size= guide_legend(title="Population"),
  color=guide_legend(title = "Fraction of GDP from industry")
  ) +
  labs(
    title = paste("Canary palm trees 2017,","n =",sum_canary$sum_2017,sep = " "),
    subtitle = "Vs. distance, population and fraction of GDP from services.",
    x = "Distance to coastline (km)",
    y = "Canary palms"
  )+ theme(
        text = element_text(size=20))
ggsave(
  "canary_palms.png",
  width = 7,
  height = 8,
  bg = "transparent"
)

predictions <- arrangeGrob(date_mean,canary_mean,nrow=2)
ggsave("predictions.png",predictions,width = 14,height=16,limitsize = FALSE,bg="transparent")
```

Using these models, we estimated the number of date and canary palms in Catalonia. Estimations appear in Figure \ref{fig:predictions} as a function of distance, population, and fraction of GDP for industry. We can see that for most species, palm trees are located closest to the coast, have a positive relationship with population and negative relationship with fraction of GDP from industry. If we consider industry to be opposed to services, it is likely this variable serves as a proxy for tourism.

\begin{figure}[!htb] 
\centerline{\includegraphics[width=30pc]{predictions.png}} 
\caption{Palm estimations according to species.}
\label{fig:predictions} 
\end{figure}

Finally, recall that these estimates correspond to 2017 estimates. To obtain 2005 estimates, we added the reported loss for date (14) and canary palms (8946) provided by GenCat. Final estimates por date and canary palms were `r sum_date$sum_2017+14` and `r sum_canary$sum_2017+8946`. 


Infested stock 2005
=====

<!-- We first define the simulation time constraints and store it in a simulation vector simtime. -->
```{r init model, include=FALSE}
START <- 2005
FINISH <- 2017
YEARS <- FINISH-START+1
STEP <- 1
simtime <- seq(START,FINISH,by=STEP) #simulation time vector
min_trees <- 0 #Min replanting
max_trees <- 100L #Max replanting 
s_Heal <- 19500L
s_Inf <- 500L
p_tau <- 0.5
```

<!-- Now, we define the stocks along with their initial values.  -->
<!-- sHeal stands for healthy trees. -->
<!-- sInf, infested trees. -->
<!-- sTre, for treated trees.  -->

```{r stocks, include=FALSE}
stocks<- c(sHeal=s_Heal, #healthy trees
           sInf=s_Inf, #infested trees
           sTre=0L #treated trees
           )
```

<!-- We now define the auxiliary variables: -->
<!-- aProb_arrival: Probability of arrival of RPW in year t.  -->
<!-- aInsfreq: Inspection frequency. -->
<!-- aTau_prob: Treatment probability. -->
<!-- aSigma_prob: Probability of succesful treatment. -->

```{r auxs, include=FALSE}
auxs <- c(
  aTau_prob = 0.8, # Treatment probability
  aSigma_prob = 0.2, #Probability of succesful treatment
  aDelay=1, #Number of years to recover from an infestation
  aEffective_Contact_Rate= 2 #Number of weevils a palm tree is exposed to per year (weevils/tree/year)
)
```

<!-- The model is a function that will take three parameters. -->

```{r model, include=FALSE }
model <-
  function(time, stocks, auxs) {
    #Three parameters for the model.
    with(as.list(c(stocks, auxs)),
         {
           #Set stocks and auxs are defined as lists.
           
           aTotal_Population <-
             sInf + sTre + sHeal #Set the total population as a sum of all three stocks.
           aBeta <-
             aEffective_Contact_Rate / aTotal_Population #Set the beta as the effective contact rate divided by the total population.
           aLambda <- aBeta * sInf
           aAffected <-  sInf + sTre
           fReplanting <-
             round(runif(1, min = min_trees, max = max_trees), 0) #Generate a random uniform.
           fFail_prev <-
             aLambda * sHeal#Define the failed prevention flow.
           fTreat <- sInf * aTau_prob#Define the treated flow.
           fUntreat <- sInf * (1 - aTau_prob)#Define untreated flow.
           fFail_treat <-
             sTre * (1 - aSigma_prob) #Define the unsuccessful treatment flow.
           fSucc_treat <-
             sTre * aSigma_prob / aDelay #Define the successfuul treatment flow.
           
           dH_dt <-
             fReplanting - fFail_prev + fSucc_treat #Define the netflow for stock healthy trees.
           dI_dt <-
             fFail_prev - fTreat - fUntreat #Define the netflow for stock infested trees.
           dTr_dt <-
             fTreat - fSucc_treat - fFail_treat #Define the netflow for stock treated trees.
        
           
           return (
             list(
               c(dH_dt, dI_dt, dTr_dt),
               ForceOfInfestation = aLambda,
               
               Treated_prob = aTau_prob,
               Success_prob = aSigma_prob,
               #Beta=aBeta,
               Treated = fTreat,
               Untreated = fUntreat,
               Failed_treatment = fFail_treat,
               Successful_treatment = fSucc_treat,
               Replanting = fReplanting,
               Total_population = aTotal_Population,
               Failed_prevention = fFail_prev,
               Affected = aAffected
             )
           )
         })
  }
o <- data.frame(ode(
  y = stocks,
  times = simtime,
  func = model,
  parms = auxs,
  method = "euler"
))#Defining integration method.
o$time <- as.integer(o$time)
o$Treated <- as.integer(o$Treated)
o$Untreated <- as.integer(o$Untreated)
o$Failed_treatment <- as.integer(o$Failed_treatment)
o$sHeal <- as.integer(o$sHeal)
o$sInf <- as.integer(o$sInf)
o$sTre <- as.integer(o$sTre)
o$Successful_treatment <- as.integer(o$Successful_treatment)
o$Total_population <- as.integer(o$Total_population)
o$Failed_prevention <- as.integer(o$Failed_prevention)
```

```{r date simulation, include=FALSE}
#Date palms model
options(scipen = 999)
CE.MIN <- 0.5
CE.MAX <- 2
tau.MIN <- 0.5
tau.MAX <- 1
sigma.MIN <- 0.01
sigma.MAX <- 0.5
sH_2005.MIN <- sum_date$sum_2017+14
sH_2005.MAX <- sum_date$sum_2017+14
sI_2005.MIN <- 1
sI_2005.MAX <- 5


parRange <- data.frame(
  min = c(CE.MIN, tau.MIN, sigma.MIN, sH_2005.MIN, sI_2005.MIN),
  max = c(CE.MAX, tau.MAX, sigma.MAX, sH_2005.MAX, sI_2005.MAX)
)

rownames(parRange) <-
  c(
    "aEffective_Contact_Rate",
    "aTau_prob",
    "aSigma_prob",
    "initHealthy",
    "initInfested"
  )


set.seed(123)
p <- data.frame(Latinhyper(parRange, 5))
g.simRuns <- list()


sensRun <- function(p) {
  g.simRuns <<- list(length = nrow(p))
  
  for (i in 1:nrow(p)) {
    initHealthy <- p [i, "initHealthy"]
    initInfested <- p [i, "initInfested"]
    auxs <- c(aDelay = 1, p[i, 1:5])
    stocks <-
      c(sHeal = initHealthy - initInfested,
        sInf = initInfested,
        sTre = 0)
    o <-
      data.frame(ode(
        y = stocks,
        simtime,
        func = model,
        parms = auxs,
        method = "euler"
      ))
    o$run <- i
    g.simRuns[[i]] <<- o
  }
}

p <- data.frame(Latinhyper(parRange, 4))
sensRun(p)
df_date <- rbind.fill(g.simRuns)

df_date %>%
mutate_at(ceiling,.vars=c("sHeal",
                         "sInf",
                        "sTre",
                        "Treated",
                        "Untreated",
                        "Failed_treatment",
                        "Successful_treatment")) -> df_date
 


```

```{r canary simulation, include=FALSE}
#Canary palms model
options(scipen = 999)
CE.MIN <- 1
CE.MAX <- 2
tau.MIN <- 0.5
tau.MAX <- 1
sigma.MIN <- 0.01
sigma.MAX <- 0.5
sH_2005.MIN <- sum_canary$sum_2017+8946
sH_2005.MAX <- sum_canary$sum_2017+8946
sI_2005.MIN <- 100
sI_2005.MAX <- 1000

parRange <- data.frame(
  min = c(CE.MIN, tau.MIN, sigma.MIN, sH_2005.MIN, sI_2005.MIN),
  max = c(CE.MAX, tau.MAX, sigma.MAX, sH_2005.MAX, sI_2005.MAX)
)

rownames(parRange) <-
  c(
    "aEffective_Contact_Rate",
    "aTau_prob",
    "aSigma_prob",
    "initHealthy",
    "initInfested"
  )


set.seed(123)
p <- data.frame(Latinhyper(parRange, 5))
g.simRuns <- list()


sensRun <- function(p) {
  g.simRuns <<- list(length = nrow(p))
  
  for (i in 1:nrow(p)) {
    initHealthy <- p [i, "initHealthy"]
    initInfested <- p [i, "initInfested"]
    auxs <- c(aDelay = 1, p[i, 1:5])
    stocks <-
      c(sHeal = initHealthy - initInfested,
        sInf = initInfested,
        sTre = 0)
    o <-
      data.frame(ode(
        y = stocks,
        simtime,
        func = model,
        parms = auxs,
        method = "euler"
      ))
    o$run <- i
    g.simRuns[[i]] <<- o
  }
}

p <- data.frame(Latinhyper(parRange, 4))
sensRun(p)
df_canary <- rbind.fill(g.simRuns)

df_canary %>%
mutate_at(ceiling,.vars=c("sHeal",
                         "sInf",
                        "sTre",
                        "Treated",
                        "Untreated",
                        "Failed_treatment",
                        "Successful_treatment")) -> df_canary
```



```{r simulations date plots, include=FALSE}
#Creating breaks for years
int_breaks <-
  function(x, n = 13)
    pretty(x, n)[pretty(x, n) %% 1 == 0]

#Simulations date
healthy_sim <- ggplot(df_date,aes(x=time,
                                    y=sHeal,
                                    group=run,
                                    fill=run,
                                    color=run))+
  geom_area(
    position = "dodge",
    alpha=0.5,linetype="dashed",size=1)+
  ylab("Healthy")+
  xlab("Time") +theme_classic()+
  scale_fill_viridis_c()+
  scale_color_viridis_c()+
  scale_x_continuous(breaks = int_breaks)+
  guides(color=FALSE,
         fill = guide_legend(title = "Run"))+
  labs(
    title = "Healthy date palms",
    x = "Time",
    y = "Date palms"
  )+theme(legend.position = "top")


early_date <- read.xlsx("early.xlsx",sheet = 3)
early_date$year <- as.double(early_date$year)
jesus_line <- ggplot(early_date,aes(x=year,y=Trees,group=Classification))+
  geom_line(color="red",
            size=1)

infested_sim <- ggplot(df_date,aes(x=time,
                                     y=sInf,
                                     fill=run,
                                     group=run,
                                     color=run))+
  geom_area(
    position = "dodge",
    alpha=0.5,linetype="dashed",size=1)+
  ylab("Infested")+
  xlab("Time") +
  theme_classic()+
  scale_fill_viridis_c()+
  scale_color_viridis_c()+
  geom_line(data=early_date,
            aes(x=year,
                y=Trees,
                group=Classification),
            color="red",
            inherit.aes = F,
            size=2,
            alpha=0.5)+
  scale_x_continuous(breaks = int_breaks)+
  guides(color=FALSE,
         fill = guide_legend(title = "Run"))+
  labs(
    title = "Infested date palms",
    subtitle = "Red line represents reported volume of infested palms.",
    x = "Time",
    y = "Date palms"
  )+theme(legend.position = "top")

treated_sim <- ggplot(df_date,aes(x=time,
                                    y=sTre,
                                    fill=run,
                                    group=run,
                                    color=run))+
  geom_area(
    position = "dodge",
    alpha=0.5,
    linetype="dashed",
    size=1)+
  ylab("Treated")+
  xlab("Time")  +
  theme_classic()+
  scale_fill_viridis_c()+
  scale_color_viridis_c()+
  scale_x_continuous(breaks = int_breaks)+
  guides(color=FALSE,
         fill = guide_legend(title = "Run"))+
  labs(
    title = "Date palms under treatment",
    x = "Time",
    y = "Date palms"
  )+theme(legend.position = "top")

simulations_date<- arrangeGrob(healthy_sim,infested_sim,treated_sim, nrow=1)
ggsave("simulations_date.png",simulations_date,width = 14,height=8,limitsize = FALSE,bg="transparent")

```

```{r simulations canary plots, include=FALSE}
#Simulations canary
healthy_sim <- ggplot(df_canary,aes(x=time,
                                    y=sHeal,
                                    group=run,
                                    fill=run,
                                    color=run))+
    geom_area(
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1)+
  ylab("Healthy")+
  xlab("Time") +theme_classic()+
  scale_fill_viridis_c()+
  scale_color_viridis_c()+
  scale_x_continuous(breaks = int_breaks)+
  guides(color=FALSE,
           fill = guide_legend(title = "Run"))+
  labs(
    title = "Healthy canary palms",
    x = "Time",
    y = "Canary palms"
  )+theme(legend.position = "top")


early_canary <- read.xlsx("early.xlsx",sheet = 2)
early_canary$year <- as.double(early_canary$year)
jesus_line <- ggplot(early_canary,aes(x=year,y=Trees,group=Classification))+
  geom_line(color="red",
            size=1)

infested_sim <- ggplot(df_canary,aes(x=time,
                                     y=sInf,
                                     fill=run,
                                     group=run,
                                     color=run))+
   geom_area(
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1)+
  ylab("Infested")+
  xlab("Time") +
  theme_classic()+
  scale_fill_viridis_c()+
    scale_color_viridis_c()+
  geom_line(data=early_canary,
            aes(x=year,
                y=Trees,
                group=Classification),
            color="red",
            inherit.aes = F,
            size=2,
            alpha=0.5)+
  scale_x_continuous(breaks = int_breaks)+
   guides(color=FALSE,
           fill = guide_legend(title = "Run"))+
  labs(
    title = "Infested Canary palms",
    subtitle = "Red line represents reported volume of infested palms.",
    x = "Time",
    y = "Canary palms"
  )+theme(legend.position = "top")

treated_sim <- ggplot(df_canary,aes(x=time,
                                    y=sTre,
                                    fill=run,
                                    group=run,
                                    color=run))+
   geom_area(
            position = "dodge",
            alpha=0.5,
            linetype="dashed",
            size=1)+
  ylab("Treated")+
  xlab("Time")  +
  theme_classic()+
  scale_fill_viridis_c()+
  scale_color_viridis_c()+
  scale_x_continuous(breaks = int_breaks)+
   guides(color=FALSE,
           fill = guide_legend(title = "Run"))+
  labs(
    title = "Canary palms under treatment",
    x = "Time",
    y = "Canary palms"
  )+theme(legend.position = "top")

simulations_canary<- arrangeGrob(healthy_sim,infested_sim,treated_sim, nrow=1)
ggsave("simulations_canary.png",simulations_canary,width = 14,height=8,limitsize = FALSE,bg="transparent")
```

```{r combine simulation plots, include=FALSE}
simulations_combined<- arrangeGrob(simulations_date,simulations_canary, nrow=2)
ggsave("simulations_combined.png",simulations_combined,width = 14,height=8,limitsize = FALSE,bg="transparent")
```

To initialise our model, we needed to estimate the number of infested trees at the beginning of the epidemic (2005). Since no information was available, we tried to estimate the number of infested trees by running several simulations of the model whilst varying the initial infested stock. As reference, we used reports for the infested stock per year. If the simulations contained the reported infested stock, then it is likely the initialisation parameters are appropriate. To run the simulations, we used Latin Hypercube Sampling (LHS) [@mckayComparisonThreeMethods1979a] implemented with the R FME package [@soetaertSolvingDifferentialEquations2010]. The reference for these estimates was government records collected by GenCat on the evolution of the pest between 2005 and 2014. According to interviews to GenCat officials, these were collected through non-systematic communications. For example, private owners or municipalities would contact GenCat and report their RPW observations. This is to say that these reports do not accurately reflect the RPW's spatio-temporal infestation pattern, but rather, they provide a minimum volume of affected palm trees per species. Four simulations with random values for each parameter are presented in Figure \ref{fig:simulations}. 

\begin{figure}[!htb]
\centerline{\includegraphics[width=30pc]{simulations_combined.png}}
\caption{Four simulations of healthy, infested and treated stocks with different parameters and initalisation values.}
\label{fig:simulations} 
\end{figure}

By looking at the infested stock for canary palms, we can see that simulation four (yellow), best fits the reports up to 2008. Afterwards, there is a decline in the reports, including a sharp decline in 2011. We believe this is due to the reporting methods and not to a sudden change in the actual patterns of infestation. This is supported by the reports from the subsequent year (2012), where reports increase and fit the third simulation. From this observation, we infer that appropiate initialization parameters for infested stock are between those of simulations three and four, which correspond to values between [3-5] for date palms and [365-975] for canary palms. These selected intialisation values are summarised along with other parameters in Table \ref{tab:initpar}. These will be used for our cost calculations. 

```{r ranges for infested stock 2005, eval=FALSE,include=FALSE}
df_canary%>%
  filter(run==3)%>%
  glimpse()
df_canary%>%
  filter(run==4)%>%
  glimpse()
#3 is run 4 - 5 is run   3 for date palms 

#365 is run 4 - 975 is run 3 for canary palms

#Filter the datasets for both species

df_date%>%
  filter(run==3|run==4)->df_date
df_canary%>%
  filter(run==3|run==4)->df_canary
```

```{r init values and parameter table, echo=FALSE}
#Canary palms model
options(scipen = 999)

Values <- c("Effective Contact Rate", 
                              "Treatment prob ($\\tau$)", 
                              "Successful treatment prob ($\\sigma$)", 
                              "Healthy stock 2005",
                              "Infested stock 2005")
Date <- c("[1, 2]","[0.5, 1]","[0.0-0.5]",(sum_date$sum_2017+14),"[3-5]")
Canary <- c("[1, 2]","[0.5, 1]","[0.0-0.5]",(sum_canary$sum_2017+8946),"[365-975]")
parameter_table <- data_frame(Values,Date,Canary)

kable(parameter_table, 
    booktabs = T, 
    caption = "\\label{tab:initpar}Initialisation and parameter values for both species.", 
    escape = FALSE) %>% 
    kable_styling(latex_options = c("hold_position"), 
        full_width = F)


```

Finally, unitary costs along with their frequencies are summarised in Table \ref{tab:unitary_costs}.
```{r unitary cost table, echo=FALSE}

Type <- as.vector(c("$c^{inspect}$",
                  "$c^{prevent}$",
                  "$c^{treat}$",
                  "$c_t^{replant}$",
                  "$c_t^{removal}$"))

`Value €`  <- as.vector(c(10,20,30,500,500))
`Yearly frequency` <- as.vector(c(2,2,5,1,1))
unit_costs <- data_frame(Type,`Value €` ,`Yearly frequency`)

kable(unit_costs,
    caption = "\\label{tab:unitary_costs}Unitary costs for management activities.", 
    booktabs = T,
    escape = F) %>% 
    kable_styling(latex_options = c("hold_position"))
```


3. Results
============

In this section we will present the calculations for management costs, benefits of palm trees and the related cost-benefit analysis (Eq. \ref{eq:nsb}). Here we present tables and figures for the aggregated results. More detailed tables and figures are included in Appendix C. 

Costs
=====

```{r exchange rates and money params, echo=FALSE,include=FALSE}
options(digits = 3)
discount_rate <- 0.05
inflation_rate <- 0.02
real_rate <- (1+discount_rate)/(1+inflation_rate)-1


USD <-  as.vector(c(12,125))
Date <- as.vector(c(2007,2014))
exchange_rate_eur <- 0.87 #quantmod::getQuote("USDEUR=X")
EUR <- USD*exchange_rate_eur #$Last
```

```{r future value function, include=FALSE,echo=FALSE}

future_value <- function(pv,real_rate,periods){

pv*(1+real_rate)^periods

}

```


As mentioned in section 2.2, costs include GenCat and management costs paid by municipalities and private palm tree owners (Eq.\ref{eq:management_costs}). The total GenCat costs were provided as a lump sum and are presented as such. Figure \ref{fig:gencat_plot} (found in Appendix C) illustrates GenCat costs brought to present value. We can see that costs are slowly rising from 2005 but in 2009 they peak and then decline with a further rise around 2014. This is due to the transfer of costs to municipalities and private owners in 2011. Before 2011, research, outreach and management costs (with the exception of replanting) are within GenCat costs, these are summarised in Appendix C (Table \ref{tab:gencat_table}) and brought to present value with a discount rate of `r discount_rate` and inflation rate of `r inflation_rate`. After 2011 private owners and municipalties assume management costs in addition to replanting.

```{r gencat costs, include=FALSE}
#Add GenCat costs. 
gov_cost <-
  c(
    30000,
    100000,
    500000,
    600000,
    1350000,
    1200000,
    475000,
    203820,
    185232,
    215110,
    140000,
    10545.15,
    8857.20
  ) 

time <- 2005:2017

gov_cost_present <- future_value(gov_cost,real_rate=real_rate,periods = (year(Sys.Date())-time))

gencat <- data_frame(time,gov_cost,gov_cost_present)

```

```{r gencast costs plots, include=FALSE}
gencat %>%
  dplyr::select(time,gov_cost_present)%>%
  gather(key="cost",
         value="amount",
         2)%>%
  ggplot(aes(x=time,
             y=amount,
             color=cost))+
  geom_area(aes(fill=cost),
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1)+
  theme_classic()+
  scale_x_continuous(breaks = int_breaks)+
  guides( color = FALSE,
           fill = FALSE)+
  labs(
    title = "GenCat costs",
    x = "Time",
    y = "Euros"
  ) + theme(legend.position = "top",
        text = element_text(size=20)) -> gencat_plot

ggsave("gencat_plot.png",gencat_plot,width = 14,height=8,limitsize = FALSE,bg="transparent")
```

We now report the costs per species and management category brought to present value. We include a lower and upper bound derived from Table \ref{tab:initpar}. Equation \ref{eq:management_costs} was applied to the lower and upper tree estimates. Tables \ref{tab:date_costs_low}, \ref{tab:date_costs_high}, \ref{tab:canary_costs_low} and \ref{tab:canary_costs_high} in Appendix C summarise each activity along with a total for management costs. The accompanying note shows an all year total. To illustrate the breakdown of costs, Figures \ref{fig:date_costs} and \ref{fig:canary_costs} present the breakdown, these are also available in Appendix C. For date palms, the costliest management activities are inspection and prevention, which makes sense since infestations for this species have been less reported. For canary palms, removal of infested trees has been the most expensive activity. This is consistent with the decline of trees and the high removal cost per tree. Table \ref{tab:total_costs} summarises the lower and upper bounds for each species in addition to total for both species (see Appendix C). This total includes GenCat costs since they are fixed in addition to date and canary costs according to lower or upper bound. An illustration of these costs is presented in Appendix C Figure \ref{fig:costs_combined_total}. The lower bound has a less steep decline than the upper bound, canary costs are higher than date costs for all years and it seems that from 2017 onwards, date costs overtake canary costs. In the upper bound were canary decline is stronger, these costs are very high but due to a reduction in overall stock, there is a declie of treatment and removal costs which allow inspection and prevention costs from date trees to become the largest cost.

```{r Management costs for date palms, include=FALSE}
#Management costs for date palms
unit_replanting <- 500 #Define the unitary cost of replanting (euros)
inspection_cost <- 10 #Define the unitary cost of an inspection (dimensionless)
prevention_cost <- 20 #Define the unitary cost of a prevention (dimensionless)
treatment_cost <- 30
removal_cost <- 500
inspection_freq <- 2 #Define the number of inspections done per year
prevention_freq <- 2 #Define the number of preventative treatments done per year
treatment_freq <- 5 #Define the number of treatments per year
#Equations

df_date %>%
  mutate(rep_cost = ifelse(time>=2011,Replanting * unit_replanting,0), #Define the replanting cost vector.  
         ins_cost = ifelse(time>=2011, sHeal * inspection_freq * inspection_cost,0),#Calculate annual inspection costs.
           pre_cost = ifelse(time>=2011, sHeal * prevention_freq * prevention_cost,0),#Calculate annual prevention costs.
        tre_cost = ifelse(time>=2011, sTre * treatment_freq * treatment_cost,0),#Calculate annual treatment costs
        rem_cost = ifelse(time>=2011, (Failed_treatment+Untreated)*removal_cost,0)#Calculate annual removal costs.
        ) %>%
  dplyr::select(run,
         time,
         rep_cost,
         ins_cost,
         pre_cost,
         tre_cost,
         rem_cost
         )-> date_costs #Calculate total management costs.
```

```{r date_costs present value,echo=FALSE}
low_run=3
date_costs%>%
  filter(run==low_run,
         time>=2011)%>%
  #Bring to present value
  mutate(rep_cost_low=future_value(rep_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         ins_cost_low=future_value(ins_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         pre_cost_low=future_value(pre_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         tre_cost_low=future_value(tre_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         rem_cost_low=future_value(rem_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         total=c(rep_cost_low+
                   ins_cost_low+
                   pre_cost_low+
                   tre_cost_low+
                   rem_cost_low))%>%
  dplyr::select(time,
                rep_cost_low,
                ins_cost_low,
                pre_cost_low,
                tre_cost_low,
                rem_cost_low,
                total)->date_cost_low


high_run=4
date_costs%>%
  filter(run==high_run,
         time>=2011)%>%
  #Bring to present value
  mutate(rep_cost_high=future_value(rep_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         ins_cost_high=future_value(ins_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         pre_cost_high=future_value(pre_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         tre_cost_high=future_value(tre_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         rem_cost_high=future_value(rem_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         total=c(rep_cost_high+
                ins_cost_high+
                pre_cost_high+
                tre_cost_high+
                rem_cost_high))%>%
  dplyr::select(time,
                rep_cost_high,
                ins_cost_high,
                pre_cost_high,
                tre_cost_high,
                rem_cost_high,
                total)->date_cost_high


```
         
```{r date costs plots, include=FALSE}

date_cost_low%>%
  gather(key = "cost",value="amount",2:6)%>%
  ggplot(
    aes(x=time,
        y=amount,
        color=cost))+
  geom_area(aes(fill=cost),
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1)+
  theme_classic()+
  scale_color_viridis_d()+
  scale_fill_viridis_d()+
  scale_x_continuous(breaks = int_breaks)+
  guides( color = FALSE,
          fill = guide_legend(title = "Cost type"))+
  labs(
    title = "Date palm costs lower bound",
    x = "Time",
    y = "Euros")+
  theme(legend.position = "top",
        text = element_text(size=20)) -> date_1

#Filter for run 6
date_cost_high%>%
  gather(key = "cost",value="amount",2:6)%>%
  ggplot(
    aes(x=time,
        y=amount,
        color=cost))+
  geom_area(aes(fill=cost),
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1)+
  theme_classic()+
  scale_color_viridis_d()+
  scale_fill_viridis_d()+
  scale_x_continuous(breaks = int_breaks)+
  guides( color = FALSE,
          fill = guide_legend(title = "Cost type"))+
  labs(
    title = "Date palm costs upper bound",
    x = "Time",
    y = "Euros")  + 
  theme(legend.position = "top",
        text = element_text(size=20))-> date_6


date_costs_plots<- arrangeGrob(date_1,date_6,nrow=2)
ggsave("date_costs.png",date_costs_plots,width = 14,height=16,limitsize = FALSE,bg="transparent")
```

```{r Management costs for canary palms, include=FALSE}
#Management costs for canary palms

#Equations
df_canary %>%
  mutate(rep_cost = ifelse(time>=2011,Replanting * unit_replanting,0), #Define the replanting cost vector.  
         ins_cost = ifelse(time>=2011, sHeal * inspection_freq * inspection_cost,0),#Calculate annual inspection costs.
         pre_cost = ifelse(time>=2011, sHeal * prevention_freq * prevention_cost,0),#Calculate annual prevention costs.
         tre_cost = ifelse(time>=2011, sTre * treatment_freq * treatment_cost,0),#Calculate annual treatment costs
         rem_cost = ifelse(time>=2011, (Failed_treatment+Untreated)*removal_cost,0)#Calculate annual removal costs.
         ) %>%
  dplyr::select(run,
         time,
         rep_cost,
         ins_cost,
         pre_cost,
         tre_cost,
         rem_cost
  )-> canary_costs #Calculate total management costs.
```

```{r canary_costs present value,echo=FALSE}

canary_costs%>%
  filter(run==3,
         time>=2011)%>%
  #Bring to present value
  mutate(rep_cost_low=future_value(rep_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         ins_cost_low=future_value(ins_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         pre_cost_low=future_value(pre_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         tre_cost_low=future_value(tre_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         rem_cost_low=future_value(rem_cost,
                                   real_rate = real_rate,
                                   periods = (year(Sys.Date())-time)),
         total=c(rep_cost_low+
                   ins_cost_low+
                   pre_cost_low+
                   tre_cost_low+
                   rem_cost_low))%>%
  dplyr::select(time,
                rep_cost_low,
                ins_cost_low,
                pre_cost_low,
                tre_cost_low,
                rem_cost_low,
                total)->canary_cost_low



canary_costs%>%
  filter(run==4,
         time>=2011)%>%
  #Bring to present value
  mutate(rep_cost_high=future_value(rep_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         ins_cost_high=future_value(ins_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         pre_cost_high=future_value(pre_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         tre_cost_high=future_value(tre_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         rem_cost_high=future_value(rem_cost,
                                    real_rate = real_rate,
                                    periods = (year(Sys.Date())-time)),
         total=c(rep_cost_high+
                   ins_cost_high+
                   pre_cost_high+
                   tre_cost_high+
                   rem_cost_high))%>%
  dplyr::select(time,
                rep_cost_high,
                ins_cost_high,
                pre_cost_high,
                tre_cost_high,
                rem_cost_high,
                total)->canary_cost_high

```

```{r canary costs plots, include=FALSE}

canary_cost_low%>%
  gather(key = "cost",value="amount",2:6)%>%
  ggplot(
    aes(x=time,
        y=amount,
        color=cost))+
  geom_area(aes(fill=cost),
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1)+
  theme_classic()+
  scale_color_viridis_d()+
  scale_fill_viridis_d()+
  scale_x_continuous(breaks = int_breaks)+
  guides( color = FALSE,
          fill = guide_legend(title = "Cost type"))+
  labs(
    title = "Canary management costs lower bound",
    x = "Time",
    y = "Euros") + theme(legend.position = "top",
        text = element_text(size=20)) -> canary_1

#Filter for run 6
canary_cost_high%>%
  gather(key = "cost",value="amount",2:6)%>%
  ggplot(
    aes(x=time,
        y=amount,
        color=cost))+
  geom_area(aes(fill=cost),
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1)+
  theme_classic()+
  scale_color_viridis_d()+
  scale_fill_viridis_d()+
  scale_x_continuous(breaks = int_breaks)+
  guides( color = FALSE,
          fill = guide_legend(title = "Cost type"))+
  labs(
    title = "Canary management costs upper bound",
    x = "Time",
    y = "Euros"
  ) + theme(legend.position = "top",
        text = element_text(size=20))-> canary_6


canary_costs_plots<- arrangeGrob(canary_1,canary_6,nrow=2)
ggsave("canary_costs.png",canary_costs_plots,width = 14,height=16,limitsize = FALSE,bg="transparent")
```

```{r Create total costs for both species including GenCat costs, include=FALSE}
#Create total costs for both species including GenCat costs

date_cost_low_vector <- as.vector(date_cost_low$total)

date_cost_low_vector <- append(date_cost_low_vector,rep(0, 6),after=0)

date_cost_high_vector  <- as.vector(date_cost_high$total)

date_cost_high_vector <- append(date_cost_high_vector,rep(0, 6),after=0)

canary_cost_low_vector  <- as.vector(canary_cost_low$total)

canary_cost_low_vector <- append(canary_cost_low_vector,rep(0, 6),after=0)

canary_cost_high_vector  <- as.vector(canary_cost_high$total)

canary_cost_high_vector <- append(canary_cost_high_vector,rep(0, 6),after=0)

total_costs <- data_frame(time,
                          gov_cost_present,
                          date_cost_low_vector,
                          date_cost_high_vector,
                          canary_cost_low_vector,
                          canary_cost_high_vector)
total_costs%>%
  mutate(total_lower =
           gov_cost_present+date_cost_low_vector+canary_cost_low_vector,
         total_upper =
           gov_cost_present+date_cost_high_vector+canary_cost_high_vector)->total_costs
names(total_costs) <- c("time",
                           "gov_cost_present",
                           "lower_date",
                           "upper_date",
                           "lower_canary",
                           "upper_canary",
                        "total_lower",
                        "total_upper")

total_costs <- total_costs%>%
  dplyr::select(time,
                gov_cost_present,
                lower_date,
                lower_canary,
                total_lower,
                upper_date,
                upper_canary,
                total_upper)
```

```{r costs_combined_total plots, include=FALSE}
total_costs %>%
  dplyr::select(time,gov_cost_present,lower_date,lower_canary)%>%
  gather(key="cost",
         value="amount",
         2:4)%>%
  ggplot(aes(x=time,
             y=amount,
             color=cost))+
  geom_area(aes(fill=cost),
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1)+
  theme_classic()+
  scale_x_continuous(breaks = int_breaks)+
  guides( color = FALSE,
           fill = guide_legend(title = "Cost type"))+
  labs(
    title = "Lower bound",
    x = "Time",
    y = "Euros"
  )+theme(legend.position = "top",
          text = element_text(size=20)) -> lower_costs_plot

total_costs %>%
  dplyr::select(time,gov_cost_present,upper_date,upper_canary)%>%
  gather(key="cost", value="amount",2:4)%>%
  ggplot(aes(x=time,
             y=amount,
             color=cost))+
  geom_area(aes(fill=cost),
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1 )+
  theme_classic()+
  scale_x_continuous(breaks = int_breaks)+
  guides( color = FALSE,
           fill = guide_legend(title = "Cost type"))+
  labs(
    title = "Upper bound",
    x = "Time",
    y = "Euros"
  )+theme(legend.position = "top",
          text = element_text(size=20)) -> upper_costs_plot

costs_combined_total<- arrangeGrob(lower_costs_plot,upper_costs_plot, nrow=2)
ggsave("costs_combined_total.png",
       costs_combined_total,
       width = 14,
       height= 16,
       limitsize = FALSE,
       bg="transparent")
```


Benefits of palm trees
========

In order to estimate the benefits associated to carbon sequestration, as described in Equation \ref{eq:benefits}, we provide an interval per tonne for the $SCC$, reflecting uncertainty about its value. A value of 12 USD was used as the lower bound since this was the lowest $SCC$ found [@iawgTechnicalUpdateSocial2016]. Similarly, the upper bound is 125 USD [@vandenberghLowerBoundSocial2014], as it was the highest found. These values were converted to euros at an exchange rate `r exchange_rate_eur` and brought to present value with a discount rate of `r discount_rate` and inflation rate of `r inflation_rate`. Table \ref{tab:scc}, summarises the original values, publication date, the values converted to euros and present value according to source, the present value. 

```{r SCC table, echo=FALSE}
PV <- future_value(USD,real_rate = real_rate,periods = (year(Sys.Date())-Date))
Source <- as.vector(c("IAWG, 2016","van den Bergh \\& Botzen, 2014"))
SCC_table <- data_frame(Source,USD,Date,EUR,PV)

SCC_table%>%
set_names(c("Source","USD","Published date","EUR", "Present value"))%>%
kable(
    caption = "\\label{tab:scc}Social cost of carbon dioxide per tonne.", 
    booktabs = T,
    escape = F) %>% 
    kable_styling(latex_options = c("hold_position"))
```


In our total benefits calculation, we will present ranges using the lowest (USD 12) and highest (USD 125) social cost of carbon $SCC$. The lowest bound for each species will be combined with lowest cost. By applying Equation \label{eq:social_cost} using a value of 0.191 tonnes per year for 1000 trees [@aguaronComparisonMethodsEstimating2012; @citeulike:14040392]. Table \ref{tab:carbon_costs} summarises benefits derived from carbon sequestration along with totals for lower and upper estimates. Since carbon sequestration values are directly proportional to the tree stock, the lower and upper bound differ according to the initial tree estimation. For the upper estimates, there is a stronger decline over time, this explains the marked difference between the 2005 and 2017 values. 

```{r  carbon sequestration benefits, include=FALSE}
#Constants
net_carbon <- 0.000191 #Define the net carbon sequestration capacity per tree (tonnes/tree)
SCC_carbon_upper <-SCC_table$PV[2] #Define the upper social cost of carbon per one tonne (euros)  
SCC_carbon_lower <- SCC_table$PV[1] #Define the upper social cost of carbon per one tonne (euros)  

date_lower <- df_date %>%
  filter(run==low_run)%>%
  mutate(total=sHeal+sInf+sTre)
  
date_upper <- df_date %>%
  filter(run==high_run)%>%
    mutate(total=sHeal+sInf+sTre)

canary_lower <- df_canary %>%
  filter(run==low_run)%>%
    mutate(total=sHeal+sInf+sTre)

canary_upper <- df_canary %>%
  filter(run==high_run)%>% 
  mutate(total=sHeal+sInf+sTre)

carbon_lower_date <- date_lower$total*SCC_carbon_lower*net_carbon
carbon_lower_date <-future_value(carbon_lower_date,
                                 real_rate = real_rate,
                                 periods = (year(Sys.Date())-time))

carbon_upper_date <- date_upper$total*SCC_carbon_upper*net_carbon
carbon_upper_date <-future_value(carbon_upper_date,
                                 real_rate = real_rate,
                                 periods = (year(Sys.Date())-time))

carbon_lower_canary <- canary_lower$total*SCC_carbon_lower*net_carbon
carbon_lower_canary <-future_value(carbon_lower_canary,
                                 real_rate = real_rate,
                                 periods = (year(Sys.Date())-time))

carbon_upper_canary <- canary_upper$total*SCC_carbon_upper*net_carbon
carbon_upper_canary <-future_value(carbon_upper_canary,
                                 real_rate = real_rate,
                                 periods = (year(Sys.Date())-time))


carbon_benefits <- data.frame(time,
                              carbon_lower_date,
                              carbon_upper_date,
                              carbon_lower_canary,
                              carbon_upper_canary)

carbon_benefits <- carbon_benefits%>%
  mutate(total_lower=carbon_lower_date+carbon_lower_canary,
         total_upper=carbon_upper_date+carbon_upper_canary)%>%
  dplyr::select(time,carbon_lower_date,carbon_lower_canary,total_lower,
                carbon_upper_date,carbon_upper_canary,total_upper)
```


Total costs and benefits 
=========

```{r total costs summary, include=FALSE}
total_costs %>%
  mutate(total_lower=gov_cost_present+lower_date+lower_canary,
         total_upper=gov_cost_present+upper_date+upper_canary) -> total_costs

carbon_benefits%>%
  mutate(total_lower=  carbon_lower_date+carbon_lower_canary,
         total_upper=carbon_lower_canary+carbon_upper_canary) -> total_benefits
```

```{r costs and benefits summary, include=FALSE}
cb_lower <- data.frame(time, total_costs$total_lower,total_benefits$total_lower)
names(cb_lower) <- c("time","costs","benefits")
cb_upper <- data.frame(time, total_costs$total_upper,total_benefits$total_lower)
names(cb_upper) <- c("time","costs","benefits")

cb_lower%>%
  summarise(sum(costs),sum(benefits)) ->cb_total_lower

cb_upper%>%
  summarise(sum(costs),sum(benefits))->cb_total_upper
```

```{r cost benefits plotted, include=FALSE}

cb_lower %>%
  mutate(costs=costs/1000)%>%
gather(key = "type",value="amount",2:3)%>%
   ggplot(aes(x=time,
             y=amount,
             color=type))+
  geom_area(aes(fill=type),
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1 )+
  theme_classic()+
  scale_x_continuous(breaks = int_breaks)+
  guides( color = FALSE,
           fill = guide_legend(title = "Cost type"))+
   scale_fill_manual(values=c("#00BA38","#F8766D"))+
  scale_color_manual(values=c("#00BA38","#F8766D"))+
  labs(
    title = "Costs vs. benefits (lower bound)",
    subtitle = "Costs were divided by 1,000 to allow benefits to appear in the plot.",
    x = "Time",
    y = "Euros"
  )+theme(legend.position = "top",
          text=element_text(size=10)) -> cb_lower_plot

cb_upper %>%
  mutate(costs=costs/1000)%>%
gather(key = "type",value="amount",2:3)%>%
   ggplot(aes(x=time,
             y=amount,
             color=type))+
  geom_area(aes(fill=type),
            position = "dodge",
            alpha=0.5,linetype="dashed",size=1 )+
  theme_classic()+
  scale_fill_manual(values=c("#00BA38","#F8766D"))+
  scale_color_manual(values=c("#00BA38","#F8766D"))+
  scale_x_continuous(breaks = int_breaks)+
  guides( color = FALSE,
           fill = guide_legend(title = "Cost type"))+
  labs(
    title = "Costs vs. benefits (upper bound)",
    subtitle = "Costs were divided by 1,000 to allow benefits to appear in the plot.",
    x = "Time",
    y = "Euros"
  )+theme(legend.position = "top",
          text=element_text(size=10)) -> cb_upper_plot

cost_benefit_plot<- arrangeGrob(cb_lower_plot,cb_upper_plot, nrow=2)
ggsave("cost_benefit_plot.png",
       cost_benefit_plot,
       width = 8,
       height= 8,
       limitsize = FALSE,
       bg="transparent")
```

By merging total costs and benefits for each bound, we obtained Figure \ref{fig:cost_benefit}. The costs include GenCat and management costs paid by municipality and private owners from 2011 onwards. The benefits associated to carbon sequestration are far exceeded by total costs.

\begin{figure}[!htb]
\centerline{\includegraphics[width=30pc]{cost_benefit_plot.png}}
\caption{Costs vs. benefits. GenCat and management costs for date and canary trees are aggregated. Benefits include carbon sequestration for both species.}
\label{fig:cost_benefit} 
\end{figure}

```{r NSB, echo=FALSE}
nsb_low <- round(sum(carbon_benefits$total_lower)-sum(total_costs$total_lower),digits = 0)
nsb_ratio_low  <- round(sum(total_costs$total_lower)/sum(carbon_benefits$total_lower),digits=0)
nsb_high <- round(sum(carbon_benefits$total_upper)-sum(total_costs$total_upper),digits=0)
nsb_ratio_high <- round(sum(total_costs$total_upper)/sum(carbon_benefits$total_upper),digits=0)
```

As stated earlier, all values presented so far have been brought to present value with the same discount (`r discount_rate`) and inflation (`r inflation_rate`) rates. By applying Equation \ref{eq:nsb} between 2005 and 2017, the net social benefits (NSB) for the lower bound are €`r nsb_low `, in other words, `r nsb_ratio_low` times benefits. For the upper bound, the NSB are €`r nsb_high ` or `r nsb_ratio_high` times the benefits. 

4. Discussion
============

```{r eval=FALSE, echo=FALSE}
wordcountaddin::text_stats(filename = "Version 5 RPW.Rmd")
```

The combined costs of pest management, eradication and replacement of RPW infested palms as well as loss of benefits were previously estimated for Italy, Spain and France. Up to 2013, the combined costs for all three countries were estimated to be €90 million, and estimations increase to €200 million by 2023 [@faoNewActionPlan2017]. Our calculations concur with magnitidue of FAO estimates although the monetary benefits and the justification for the conservation of palm trees was not found. 

Places where a similar cost benefit analysis could yield results that have an overall positive effect on society is for example, Tunisia, with an economy that largely depends on date exports and allegedly the RPW is a major threat [@speakmancordallPanicRedDevils2017]. Another case where this type of conservation policy could be justified is the Canary Islands since canary palms are a native species, unlike in the Iberian Peninsula. When a species is native to an area, conservation efforts from a socio-cultural perspective are more likely to be justified. In these islands, the RPW was detected in 2007 and declared completely eradicated by 2016. Eradication efforts included the inspection of 706,081 palms, of which 209,537 were treated, 659 removed and destroyed, and 681 weevils were caught [@gobiernodecanariasCanariasPrimerTerritorio2016].

The larger mandate that underpins the conservation policy here analysed is 2007/365/CE (Appendix). This policy was released for EU member states with susceptible trees stating the measures to be taken, including the ones here described as management acitivites. No associated cost-benefit analysis was found justifying the EU wide mandate. Prioritisation of budgets for biological invasions should be based on analysis that include species, pathways and sites, and impacts [@mcgeochPrioritizingSpeciesPathways2016]. 

Understandably, an argument against the type of cost-benefit analysis here performed is that social and cultural values are hard to capture and quantify. Granted, non-monetary values are important, as we acknowledged by including the TEV framework. In a world with unlimited resources, conservation of natural assets without any direct or indirect benefits, could be possible. Alas, when dealing with biological insect invasions that critically affect human health, livelihoods, threatened species, etc. priortising the cosliest impacts is necessary. We suggest the EU should perhaps reconsider this policy and similar ones to allow each member state to perform its own analysis and decide if these types of policies are locally justified.

5. Conclusions
============

In the context of rising biological invasions with costly impacts, determining areas of priority is fundamental for optimal budgeting. Several frameworks for prioritising these type of invasions have risen in recent years. In this article we assess to what extent these frameworks are being implemented in current policies. To do so, we selected a policy designed in response to an invasive insect, the Red Palm Weevil (RPW). Amongst invasive species, insects are the costliest and it is estimated less than half of potential insect invasions have taken place. To assess the policy, we performed a cost-benefit analysis within the region of Catalonia, Spain. This was done in several steps. First, we interviewed stakeholders from goverment and industry to properly understand the local reality of the RPW infestation. Second, we estimated the number of palm trees in the region at the beginning of the infestation and modelled it until present. Third, we calculated the costs and benefits associated with the policy. Finally, we estimated the net social benefits of the policy. We found that the costs of the policy far outweight the benefits. This is likely due to the almost exclusive ornamental use of palm trees. 



Acknowledgements
==========

Servei de Sanitat Vegetal de la Generalitat de Catalunya provided interviews, records and further information on the pest in Catalonia. Dr Martí Boada and Dr Jaume Marlès Magre kindly gave feedback on the estimation of the number of palm trees. Joan Freixa Gaspar and Josep Muntaner shared insights on private palm tree owners and their behaviour. Cristian E. Nuno from the University of Chicago Urban Labs provided guidance on calculating distances to the coastline. 


Funding
==========

This research was funded by the "María de Maeztu" Unit of Excellence 2016 from the Spanish Ministry of Economy and Competitiveness (MINECO), project number MDM-2015-0552.

\newpage

Appendices
==========


Appendix A: Summaries of the policies selected and related legislation
========

This section extends on relevant details on the policy assessed in this article. It summarises all GenCat orders and includes the EU legislation that underpins RPW related legislation. 

-   Order ARP/343/2006: On July 3rd 2006, the presence of the RPW in Catalonia is confirmed. Pest prevention and management is declared in the public’s interest. This declaration is done based on Article 15 from Spanish Law 43/2002 which states that when a potential pest (i.e. economic or environmental impacts) is identified for the first time it should be immediately declared. The management against this pest is considered in the public’s utility or greater good if it meets several criteria, including that it is present in areas that were previously declared pest free [INSERT HERE BOE citation.] Costing implications based on articles 8, 9 and 12 of this order include prospection, detection, destruction and indemnification to property owners begin. The total costs for these activities have been provided by the regional government, which also include survey, research, regulation, management, and outreach costs [citation]. 

-   2007/365/CE: On 25th May 2007, the EU Commission published a decision based on previous legislation for plant protection and the RPW spread along the Iberian Peninsula. This decision specifies the vulnerability of palm trees and announces a series of measures to be taken by member countries, including the delineation of areas that are in danger. 

-   2008/776/CE: On 6th October 2008, a subsequent decision was reached by the EU where the 2007 decision is modified to expand the range of palm trees vulnerable to the RPW.

-   Order AAR/226/2009: Based on the EU commission, the GenCat updates the 2006 order by redefining and specifying actions to be taken, namely delimitation, inspections, pruning, treatment, and destruction of affected palm trees. It restates that all costs continue to be borne by the GenCat.

-   2010/467/UE: On 28th August 2010, a correction to the 17th August 2010 decision is published that states that if routine inspections from the past three years show that eradication seems to not be feasible within a year, then local measures should focus on pest contention and suppression within the selected area but maintaining eradication as a long-term objective. 

-   Order AAR/2802/2010 and Resolution AAR/2802/2010: On 20th October 2010, a delimited area for the pest is published in the form of municipalities. 

-   Order AAM/56/2011: On April 12th 2011, a modification to the 2006 order was published. The main changes are done to articles 8, 9, 11, 12. The key change made from this final order is that GenCat from this year onwards continues to perform annual inspections, publish the delimited area and inform local authorities through outreach activities, but treatment and removal costs are transferred to municipalities and private owners. These activities transferred are mandatory and procedure are carefully described including the type of chemicals to be used, times of the year when they should be done and specific places where trees should be destroyed. 

Appendix B: Variable symbols
========
For reference, this appendix summarises variable symbols for stocks (\ref{tab:stocks}) and flows (\ref{tab:flows}). These were not included in the text to avoid repetition although some readers might find these tables helpful.

```{r stock table summary, echo=FALSE}
Summary_Stock <- as.vector(c("Healthy","Infested","In treatment","Lost","Replanted","Total population"))
Stock <- as.vector(c("$S_t^H$","$S_t^I$","$S_t^{T}$","$S_t^L$","$S_t^R$","$S_t$"))
df_stock <- as_tibble(Stock,Summary_Stock)

colnames(df_stock) <- c("Stock")
kable(df_stock, caption = "\\label{tab:stocks}Palm tree stocks.", booktabs = T,escape = F) %>%
kable_styling(latex_options = c("hold_position"))
```

```{r flows table summary, echo=FALSE}

Summary_Flow <- as.vector(c("Replanted",
                            "Failed prevention",
                            "Untreated",
                            "Treated",
                            "Failed treatment",
                            "Successful treatment"))
Flow <- as.vector(c("$F_{t,t-1}^R$",
                    "$F_{t,t-1}^{FP}$",
                    "$F_{t,t-1}^{UTR}$",
                    "$F_{t,t-1}^{TR}$",
                    "$F_{t,t-1}^{FT}$",
                    "$F_{t,t-1}^{ST}$"
                    ))	

df_flow <- as_tibble(Flow,Summary_Flow)
colnames(df_flow) <- c("Flow")
kable(df_flow, caption = "\\label{tab:flows}Palm tree flows.", booktabs = T,escape = F) %>%
kable_styling(latex_options = c("hold_position"))
```

Appendix C: Detailed results 
========

This section expands on the result section by providing the details of each element that contributes to the costs and benefits. The folowing tables and figures are included in the text but aggregated in the total costs and benefits. 

```{r gencat cost tables, echo=FALSE}
gencat%>%
  set_names(c("Year","GenCat cost","GenCat cost present"))%>%
kable(
  caption = "\\label{tab:gencat_table}GenCat costs associated to RPW expenditure over time.",
  booktabs = T,
  escape = F,
  linesep= ""
) %>%
  kable_styling(latex_options = c("hold_position"))%>%
  footnote(general= c(
                      paste("Total cost of all activities for all years is ",
                         round(sum(gencat$gov_cost_present),
                               digits = 0),
                         ".",sep = ""),
                      "Source: GENCAT, 2018."),
                     footnote_as_chunk = T)
```

\begin{figure}[!htb]
\centerline{\includegraphics[width=30pc]{gencat_plot.png}}
\caption{GenCat costs overtime associated to RPW expenditure.}
\label{fig:gencat_plot} 
\end{figure}

```{r date_cost_low table}
date_cost_low%>%
  set_names(c("Year","Replanting","Inspection","Prevention","Treatment","Removal","Total"))%>%
  kable(
    caption = "\\label{tab:date_costs_low}Management costs (lower bound) for date palms over time.",
    booktabs = T,
    escape = F,
    linesep= ""
  ) %>%
  kable_styling(latex_options = c("hold_position"))%>%
  footnote(general=paste("Total cost of all activities for all years is ",round(sum(date_cost_low$total),digits = 0),".",sep = ""),
           footnote_as_chunk = T)
```

```{r date_cost_high table}
date_cost_high%>%
  set_names(c("Year","Replanting","Inspection","Prevention","Treatment","Removal","Total"))%>%
  kable(
    caption = "\\label{tab:date_costs_high}Management costs (upper bound) for date palms over time.",
    booktabs = T,
    escape = F,
    linesep= ""
  ) %>%
  kable_styling(latex_options = c("hold_position"))%>%
   footnote(general=paste("Total cost of all activities for all years is ",round(sum(date_cost_high$total),digits = 0),".",sep = ""),
           footnote_as_chunk = T)


```

\begin{figure}[!htb]
\centerline{\includegraphics[width=30pc]{date_costs.png}}
\caption{Management costs over time for date palms by activity.}
\label{fig:date_costs} 
\end{figure}

```{r canary_cost_low table}
canary_cost_low%>%
  set_names(c("Year","Replanting","Inspection","Prevention","Treatment","Removal","Total"))%>%
  kable(
    caption = "\\label{tab:canary_costs_low}Management costs (lower bound) for canary palms over time.",
    booktabs = T,
    escape = F,
    linesep= ""
  ) %>%
  kable_styling(latex_options = c("hold_position"))%>%
   footnote(general=paste("Total cost of all activities for all years is ",round(sum(canary_cost_low$total),digits = 0),".",sep = ""), footnote_as_chunk = T)
```

```{r canary_cost_high table}
canary_cost_high%>%
  set_names(c("Year","Replanting","Inspection","Prevention","Treatment","Removal","Total"))%>%
  kable(
    caption = "\\label{tab:canary_costs_high}Management costs (upper bound) for canary palms over time.",
    booktabs = T,
    escape = F,
    linesep= ""
  ) %>%
  kable_styling(latex_options = c("hold_position"))%>%
   footnote(general=paste("Total cost of all activities for all years is ",round(sum(canary_cost_high$total),digits = 0),".",sep = ""), footnote_as_chunk = T)
```

\begin{figure}[!htb]
\centerline{\includegraphics[width=30pc]{canary_costs.png}}
\caption{Management costs over time for canary palms by activity.}
\label{fig:canary_costs} 
\end{figure}

```{r total costs table, echo=FALSE}
total_costs%>%
   set_names(c("Year","GenCat","Date","Canary","Total","Date","Canary","Total"))%>%
  kable(
    caption = "\\label{tab:total_costs}Total costs.",
    booktabs = T,
    escape = F,
    linesep= ""
  ) %>%
  kable_styling(latex_options = c("hold_position"))%>%
  add_header_above(c("","","Lower"=3,"Higher"=3))%>%
  footnote(general=paste("Total for lower bound is ",round(sum(total_costs$total_lower),digits=0),", total for higher bound is ",round(sum(total_costs$total_upper),digits=0),sep = ""), footnote_as_chunk = T)
```

\begin{figure}[!htb]
\centerline{\includegraphics[width=30pc]{costs_combined_total.png}}
\caption{Total costs over time for GenCat, date and canary palms.}
\label{fig:costs_combined_total} 
\end{figure}

```{r carbon benefits table}
carbon_benefits%>%
  set_names(c("Year","Date","Canary","Total","Date","Canary","Total"))%>%
kable(
    caption = "\\label{tab:carbon_costs}Carbon sequestration values over time.",
    booktabs = T,
    escape = F,
    linesep= ""
  ) %>%
  kable_styling(latex_options = c("hold_position"))%>%
  add_header_above(c("","Lower"=3,"Higher"=3))%>%
  footnote(general=paste("Combining the totals for all years for both species yields a lower bound of ",round(sum(carbon_benefits$total_lower),digits=0)," and a higher bound of ",round(sum(carbon_benefits$total_upper),digits=0),".",sep = ""),
           footnote_as_chunk = T)

```

\pagebreak
\setlength{\parindent}{-0.5in}
      \setlength{\leftskip}{0.5in}
      \setlength{\parskip}{8pt}
      
References {#references .unnumbered}
==========

